---
title: "Permutation-Assisted Group Lasso: A Simulation Study"
subtitle: "Comparing early and late fusion strategies for multi-modal feature selection"
author: "Anurag Nepal"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Background

I wanted a fair way to compare feature selection strategies when features arrive in correlated blocks from different sources. Plain lasso ignores group structure. Group lasso respects it, but it usually penalizes every group the same way. I test whether data-driven penalties and a permutation-based selection rule do better. The question this study answers: which combination of fusion strategy and penalty recovers the true features with the fewest false picks?

## Setup

The simulation builds a dataset where I know the ground truth. There are 100 samples, three causal blocks of 10 correlated features each, and 200 pure noise features. The binary outcome comes from a logistic model on the latent causal variables, with intercept -1 and effects 1.5, -2 and 1.2. I split the features into two mock modalities: modality A holds blocks 1 and 2 plus half the noise, modality B holds block 3 plus the rest. I repeat everything over 5 runs with different seeds.

## Method

First, the packages.

```{r libraries}
# Data manipulation and visualization
library(tidyverse)
library(pheatmap)
library(knitr)

# Modeling and evaluation
library(glmnet)
library(grpreg)
library(pROC)

# Adaptive weights calculation
library(SKAT)
library(Matrix)
```

The data generator. Each causal block starts from one latent uniform variable. The ten features in a block are noisy copies of it, with the noise level rising across the block so the features are correlated but not identical. The outcome is drawn from a logistic model on the latent variables.

```{r simulate-data}
#' @title Simulate Data with Correlated Features and a Binary Outcome
simulation_data <- function(EF = c(-1, 1.5, -2, 1.2, -1.8, 1.0),
                              n_ss = 250,
                              n_x = rep(10, 5),
                              n_noise = 200,
                              SD_v = 0.5) {

  num_causal <- length(EF) - 1
  x <- matrix(runif(n_ss * num_causal), nrow = n_ss, ncol = num_causal)
  colnames(x) <- paste0("x", 1:num_causal)

  v <- list()
  for (i in 1:num_causal) {
    v_i <- matrix(NA, nrow = n_ss, ncol = n_x[i])
    for (j in 1:n_x[i]) {
      noise <- rnorm(n = n_ss, mean = 0, sd = SD_v)
      v_i[, j] <- x[, i] + (0.01 + 0.5 * (j - 1) / (n_x[i] - 1)) * noise
    }
    colnames(v_i) <- paste0("v", i, "_", 1:n_x[i])
    v[[i]] <- v_i
  }
  v_df <- do.call(cbind, v)

  w <- matrix(runif(n_ss * n_noise), nrow = n_ss, ncol = n_noise)
  colnames(w) <- paste0("w", 1:n_noise)

  linear_predictor <- EF[1] + x %*% EF[-1]
  prob <- 1 / (1 + exp(-linear_predictor))
  y <- rbinom(n_ss, 1, prob)

  sim_df <- as.data.frame(cbind(v_df, w))
  sim_df$y <- y
  sim_df <- sim_df[, c("y", colnames(v_df), colnames(w))]

  return(sim_df)
}
```

Groups are not given to the method. They are learned by hierarchical clustering on 1 minus the absolute Spearman correlation, cut so there is roughly one group per eight features.

```{r define-groups}
#' @title Define Groups via Clustering
define_groups <- function(feature_df) {
  cor_matrix <- cor(feature_df, method = "spearman")
  dist_matrix <- as.dist(1 - abs(cor_matrix))
  hclust_obj <- hclust(dist_matrix, method = "ward.D2")

  num_groups <- max(1, round(ncol(feature_df) / 8))
  feature_groups_vec <- cutree(hclust_obj, k = num_groups)

  return(list(groups = feature_groups_vec, hclust = hclust_obj, cor = cor_matrix))
}
```

Two ways to set the group penalties. The standard version ranks groups by their mean within-group correlation and penalizes loosely correlated groups more. The adaptive version runs a SKAT test per group and converts the p-value into a penalty of 1 / (-log(p) + 1). Groups with strong evidence get penalized less.

```{r penalty-functions}
#' @title Calculate Penalties based on Correlation (Standard)
calculate_correlation_penalties <- function(feature_df, groups, cor_matrix) {

  unique_groups <- unique(groups)
  group_info <- data.frame(feature = names(groups), group = groups)

  mean_cor_per_group <- sapply(unique_groups, function(g) {
    members <- group_info$feature[group_info$group == g]
    if (length(members) < 2) return(0)
    sub_cor_matrix <- cor_matrix[members, members]
    mean(abs(sub_cor_matrix[upper.tri(sub_cor_matrix)]))
  })

  group_summary <- data.frame(group = unique_groups, mean_cor = mean_cor_per_group) %>%
    left_join(group_info %>% dplyr::count(group, name = "size"), by = "group") %>%
    mutate(
      importance_rank = rank(-mean_cor, ties.method = "min"),
      penalty_raw = importance_rank * sqrt(size)
    )

  penalties <- group_summary$penalty_raw / mean(group_summary$penalty_raw)
  names(penalties) <- group_summary$group

  return(penalties[as.character(sort(unique(groups)))])
}

#' @title Calculate Penalties based on SKAT P-values (Adaptive)
calculate_skat_penalties <- function(X, y, groups) {

  # 1. Fit null model
  obj <- SKAT_Null_Model(y ~ 1, out_type = "D")

  unique_groups <- sort(unique(groups))
  p_values <- numeric(length(unique_groups))
  names(p_values) <- unique_groups
  group_sizes <- numeric(length(unique_groups))

  X_mat <- as.matrix(X)

  for (g in unique_groups) {
    g_features <- names(groups)[groups == g]
    Z <- X_mat[, g_features, drop = FALSE]
    group_sizes[g] <- length(g_features)

    tryCatch({
      suppressWarnings({
        skat_res <- SKATBinary(Z, obj, kernel = "Linear.Weighted")
      })
      p_values[g] <- skat_res$p.value
    }, error = function(e) {
      p_values[g] <<- 1
    })
  }

  # 2. Calculate penalties: sqrt(size) * [ 1 / (-log(p) + 1) ]
  epsilon <- 1e-6
  neg_log_p <- -log(p_values + epsilon)
  neg_log_p[neg_log_p < 0] <- 0
  raw_penalty <- sqrt(group_sizes) * (1 / (neg_log_p + 1))

  # 3. Normalize
  final_penalty <- raw_penalty / mean(raw_penalty)

  return(final_penalty)
}
```

The selection rule. For each of 25 permutations I shuffle the rows to make a decoy copy of every feature, fit a group lasso path on the real plus decoy features, and pick the lambda where the real selections beat the decoy selections by the most. A feature is kept if it is selected in at least half the permutations.

```{r perm-assisted-selection}
#' @title Permutation-Assisted Group Lasso
perm_assisted_grplasso_binary <- function(X, y, n_perm, feature_groups, group_penalties, cutoff) {
  X <- as.matrix(X)
  p <- ncol(X)
  all_feature_names <- colnames(X)
  selected_features_all_perms <- c()

  knockoff_groups <- feature_groups + max(feature_groups)
  full_group_penalties <- c(group_penalties, group_penalties)

  for (k in 1:n_perm) {
    X_knockoff <- X[sample(nrow(X)), ]

    fit <- grpreg(
      cbind(X, X_knockoff), y,
      group = c(feature_groups, knockoff_groups),
      penalty = "grLasso",
      family = "binomial",
      group.multiplier = full_group_penalties,
      nlambda = 50,
      lambda.min = 0.05
    )

    betas <- as.matrix(fit$beta[-1, ])

    n_original <- colSums(abs(betas[1:p, ]) > 0)
    n_knockoff <- colSums(abs(betas[(p + 1):(2 * p), ]) > 0)
    diff <- n_original - n_knockoff

    if (all(diff <= 0)) {
        optimal_idx <- NULL
    } else {
        optimal_idx <- which.max(diff)
    }

    if (!is.null(optimal_idx)) {
      selected_indices <- which(abs(betas[1:p, optimal_idx]) > 0)
      if (length(selected_indices) > 0) {
        selected_features_all_perms <- c(selected_features_all_perms, rownames(betas)[selected_indices])
      }
    }
  }

  freq_counts <- setNames(rep(0, p), all_feature_names)
  if (length(selected_features_all_perms) > 0) {
    counts_table <- table(selected_features_all_perms)
    freq_counts[names(counts_table)] <- counts_table
  }

  feature_frequencies <- freq_counts / n_perm
  final_selection <- names(feature_frequencies)[feature_frequencies >= cutoff]

  return(list(final_selection = final_selection, feature_frequencies = feature_frequencies))
}
```

Scoring is against the known causal features: F1, sensitivity, false discovery rate, and the number of features selected. AUC comes from a refit on the selected features. I also keep a Jaccard helper for selection stability across runs.

```{r metric-helpers}
calculate_metrics <- function(selected, true) {
  if (length(selected) == 0) {
    tp <- 0; fp <- 0
  } else {
    tp <- sum(selected %in% true)
    fp <- length(selected) - tp
  }
  fn <- length(true) - tp

  precision <- ifelse((tp + fp) == 0, 0, tp / (tp + fp))
  sensitivity <- ifelse((tp + fn) == 0, 0, tp / (tp + fn))
  f1_score <- ifelse((precision + sensitivity) == 0, 0,
                     2 * (precision * sensitivity) / (precision + sensitivity))
  fdr <- ifelse((tp + fp) == 0, 0, fp / (tp + fp))

  return(data.frame(F1_Score = f1_score, Sensitivity = sensitivity, FDR = fdr, Num_Selected = length(selected)))
}

calculate_subsequent_jaccard <- function(feature_list) {
  n <- length(feature_list)
  if (n < 2) return(NA)
  jaccard_scores <- numeric(n - 1)
  for (i in 1:(n - 1)) {
    set1 <- feature_list[[i]]; set2 <- feature_list[[i+1]]
    intersection <- length(intersect(set1, set2))
    union <- length(union(set1, set2))
    jaccard_scores[i] <- ifelse(union == 0, 1, intersection / union)
  }
  return(mean(jaccard_scores))
}
```

Now the comparison. Five methods: early fusion with standard penalties, early fusion with adaptive penalties, late fusion with standard penalties, late fusion with adaptive penalties, and a baseline lasso. This chunk takes a while, so its console chatter is hidden. The summary table is built at the end.

```{r run-comparison, results='hide'}
# --- MAIN SIMULATION LOOP ---

N_FEATURES_PER_BLOCK <- rep(10, 3)
BASE_INTERCEPT <- -1
BASE_COEFFICIENTS <- c(1.5, -2, 1.2)

# Parameters: one scenario (n = 100, full effects, 200 noise features)
param_n_samples <- c(100)
param_effect_multipliers <- c(1)
param_n_noise <- c(200)

N_RUNS <- 5
N_PERMUTATIONS <- 25
SELECTION_CUTOFF <- 0.5

all_experiment_metrics <- list()
run_counter <- 1

for (n_val in param_n_samples) {
  for (ef_mult in param_effect_multipliers) {
    for (n_noise_val in param_n_noise) {

      TRUE_COEFFICIENTS <- c(BASE_INTERCEPT, BASE_COEFFICIENTS * ef_mult)

      run_metrics_list <- list()

      for (i in 1:N_RUNS) {
        set.seed(i)

        # 1. Data generation
        full_data <- simulation_data(EF = TRUE_COEFFICIENTS, n_ss = n_val,
                                     n_x = N_FEATURES_PER_BLOCK, n_noise = n_noise_val)
        y <- full_data$y

        # 2. Define features and mock modalities
        true_signal_features <- colnames(full_data)[grepl("^v[1-3]_", colnames(full_data))]
        modA_features <- c(colnames(full_data)[grepl("^v[1-2]_", colnames(full_data))],
                           colnames(full_data)[grepl("^w", colnames(full_data))][1:floor(n_noise_val/2)])
        modB_features <- c(colnames(full_data)[grepl("^v3_", colnames(full_data))],
                           colnames(full_data)[grepl("^w", colnames(full_data))][(floor(n_noise_val/2)+1):n_noise_val])

        X_all <- full_data[, c(modA_features, modB_features)]
        X_all_scaled <- scale(as.matrix(X_all))
        X_modA_scaled <- scale(as.matrix(full_data[, modA_features]))
        X_modB_scaled <- scale(as.matrix(full_data[, modB_features]))

        # 3. Groups and penalties
        grp_struct_early <- define_groups(X_all)
        pen_std_early <- calculate_correlation_penalties(X_all, grp_struct_early$groups, grp_struct_early$cor)
        pen_adp_early <- calculate_skat_penalties(X_all, y, grp_struct_early$groups)

        grp_struct_A <- define_groups(full_data[, modA_features])
        pen_std_A <- calculate_correlation_penalties(full_data[, modA_features], grp_struct_A$groups, grp_struct_A$cor)
        pen_adp_A <- calculate_skat_penalties(full_data[, modA_features], y, grp_struct_A$groups)

        grp_struct_B <- define_groups(full_data[, modB_features])
        pen_std_B <- calculate_correlation_penalties(full_data[, modB_features], grp_struct_B$groups, grp_struct_B$cor)
        pen_adp_B <- calculate_skat_penalties(full_data[, modB_features], y, grp_struct_B$groups)

        # 4. Run the five methods
        res_early_std <- perm_assisted_grplasso_binary(
          X_all_scaled, y, N_PERMUTATIONS, grp_struct_early$groups, pen_std_early, SELECTION_CUTOFF
        )
        res_early_adp <- perm_assisted_grplasso_binary(
          X_all_scaled, y, N_PERMUTATIONS, grp_struct_early$groups, pen_adp_early, SELECTION_CUTOFF
        )

        res_late_std_A <- perm_assisted_grplasso_binary(X_modA_scaled, y, N_PERMUTATIONS, grp_struct_A$groups, pen_std_A, SELECTION_CUTOFF)
        res_late_std_B <- perm_assisted_grplasso_binary(X_modB_scaled, y, N_PERMUTATIONS, grp_struct_B$groups, pen_std_B, SELECTION_CUTOFF)
        sel_late_std <- unique(c(res_late_std_A$final_selection, res_late_std_B$final_selection))

        res_late_adp_A <- perm_assisted_grplasso_binary(X_modA_scaled, y, N_PERMUTATIONS, grp_struct_A$groups, pen_adp_A, SELECTION_CUTOFF)
        res_late_adp_B <- perm_assisted_grplasso_binary(X_modB_scaled, y, N_PERMUTATIONS, grp_struct_B$groups, pen_adp_B, SELECTION_CUTOFF)
        sel_late_adp <- unique(c(res_late_adp_A$final_selection, res_late_adp_B$final_selection))

        cv_fit_lasso <- cv.glmnet(X_all_scaled, y, family = "binomial", alpha = 1)
        lasso_coefs <- coef(cv_fit_lasso, s = "lambda.min")
        sel_lasso <- rownames(lasso_coefs)[which(lasso_coefs != 0)]
        sel_lasso <- sel_lasso[sel_lasso != "(Intercept)"]

        # 5. AUC from a refit on the selected features
        calc_auc <- function(X_mat, sel_vars, y_target) {
          if (length(sel_vars) == 0) return(0.5)
          if (length(sel_vars) == 1) {
             df_single <- data.frame(y = y_target, x = X_mat[, sel_vars])
             fit_glm <- glm(y ~ x, data = df_single, family = "binomial")
             preds <- predict(fit_glm, type = "response")
             return(as.numeric(auc(roc(y_target, as.vector(preds), quiet = TRUE))))
          }
          fit_cv <- cv.glmnet(X_mat[, sel_vars, drop = FALSE], y_target, family = "binomial")
          preds <- predict(fit_cv, newx = X_mat[, sel_vars, drop = FALSE], s = "lambda.min", type = "response")
          return(as.numeric(auc(roc(y_target, as.vector(preds), quiet = TRUE))))
        }

        auc_early_std <- calc_auc(X_all_scaled, res_early_std$final_selection, y)
        auc_early_adp <- calc_auc(X_all_scaled, res_early_adp$final_selection, y)
        auc_lasso     <- calc_auc(X_all_scaled, sel_lasso, y)

        get_probs <- function(X_mat, sel_vars, y_target, n) {
          if (length(sel_vars) == 0) return(rep(0.5, n))
          if (length(sel_vars) == 1) {
             df_single <- data.frame(y = y_target, x = X_mat[, sel_vars])
             fit_glm <- glm(y ~ x, data = df_single, family = "binomial")
             return(predict(fit_glm, type = "response"))
          }
          fit <- cv.glmnet(X_mat[, sel_vars, drop = FALSE], y_target, family = "binomial")
          return(predict(fit, newx = X_mat[, sel_vars, drop = FALSE], s = "lambda.min", type = "response"))
        }

        probs_std_A <- get_probs(X_modA_scaled, res_late_std_A$final_selection, y, n_val)
        probs_std_B <- get_probs(X_modB_scaled, res_late_std_B$final_selection, y, n_val)
        auc_late_std <- as.numeric(auc(roc(y, as.vector((probs_std_A + probs_std_B) / 2), quiet = TRUE)))

        probs_adp_A <- get_probs(X_modA_scaled, res_late_adp_A$final_selection, y, n_val)
        probs_adp_B <- get_probs(X_modB_scaled, res_late_adp_B$final_selection, y, n_val)
        auc_late_adp <- as.numeric(auc(roc(y, as.vector((probs_adp_A + probs_adp_B) / 2), quiet = TRUE)))

        # 6. Score against the known truth
        run_metrics_list[[i]] <- bind_rows(
          data.frame(Model = "Early Fusion (Std)", AUC = auc_early_std, calculate_metrics(res_early_std$final_selection, true_signal_features)),
          data.frame(Model = "Early Fusion (Adap)", AUC = auc_early_adp, calculate_metrics(res_early_adp$final_selection, true_signal_features)),
          data.frame(Model = "Late Fusion (Std)", AUC = auc_late_std, calculate_metrics(sel_late_std, true_signal_features)),
          data.frame(Model = "Late Fusion (Adap)", AUC = auc_late_adp, calculate_metrics(sel_late_adp, true_signal_features)),
          data.frame(Model = "Baseline LASSO", AUC = auc_lasso, calculate_metrics(sel_lasso, true_signal_features))
        )
      }

      summary_metrics <- do.call(rbind, run_metrics_list) %>%
        group_by(Model) %>%
        summarise(across(c(AUC, F1_Score, Sensitivity, FDR, Num_Selected), list(Mean = mean, SD = sd))) %>%
        mutate(n_samples = n_val, effect_multiplier = ef_mult, n_noise = n_noise_val)
      all_experiment_metrics[[run_counter]] <- summary_metrics

      run_counter <- run_counter + 1
    }
  }
}

final_metrics_df <- do.call(rbind, all_experiment_metrics)
```

## Results

Mean and standard deviation of each metric across the 5 runs. F1 and sensitivity are scored against the 30 known causal features.

```{r results-table}
knitr::kable(final_metrics_df, digits = 3,
             caption = "Method comparison: mean (SD) over 5 simulation runs, n = 100")
```

F1 scores at a glance, with one-SD bars:

```{r results-plot, fig.width=7, fig.height=4}
ggplot(final_metrics_df,
       aes(x = reorder(Model, F1_Score_Mean), y = F1_Score_Mean, fill = Model)) +
  geom_col() +
  geom_errorbar(aes(ymin = F1_Score_Mean - F1_Score_SD,
                    ymax = F1_Score_Mean + F1_Score_SD), width = 0.2) +
  coord_flip() +
  labs(title = "F1 score by method (mean and SD over 5 runs)",
       x = NULL, y = "F1") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Takeaway

This study is built so the table above answers the question directly. If the adaptive penalties work, the adaptive rows should show higher F1 and lower false discovery rates than their standard counterparts at a similar AUC. The late fusion rows test whether modeling each modality on its own and combining the picks beats throwing everything into one joint model. Because the truth is known here, a win in this table means the method itself is sound before it ever touches real data.
