---
title: "Multi-Modal Feature Selection on TCGA Breast Cancer Data"
subtitle: "Adaptive group lasso with permutation-based penalties on RNA-seq and methylation"
author: "Anurag Nepal"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Background

Simulations are clean and real data is not. I apply the same adaptive group lasso idea to TCGA breast cancer data with two real modalities, RNA sequencing and DNA methylation, predicting vital status. The question this analysis answers: does the method pick sensible features on real multi-modal data, and are those picks stable when the data shifts a little?

## Setup

The data come from the TCGA-BRCA cohort through the curatedTCGAData Bioconductor package: the RNASeq2GeneNorm assay and the methylation assays, version 2.1.1. Patients are matched across the two modalities by the first 12 characters of their barcodes. The outcome is vital status, coded 1 for deceased and 0 for living. Each modality is cleaned (features that are all NA are dropped, remaining NAs are imputed with row means) and filtered down to the 1000 highest-variance features.

## Method

First, the packages. The TCGA download needs curatedTCGAData, TCGAutils and MultiAssayExperiment from Bioconductor.

```{r libraries}
# --- Data manipulation and viz ---
library(tidyverse)
library(pheatmap)
library(knitr)
library(ggplot2)

# --- Modeling ---
library(glmnet)
library(grpreg)
library(pROC)

# --- TCGA data access ---
# If missing: BiocManager::install(c("curatedTCGAData", "TCGAutils", "MultiAssayExperiment"))
library(curatedTCGAData)
library(TCGAutils)
library(MultiAssayExperiment)
```

The data loader. It fetches both assays, forces the results into plain numeric matrices (the raw objects can be DelayedArray or HDF5-backed), matches patients across modalities, pulls vital status from the clinical table, and keeps the top 1000 variable features per modality.

```{r load-data-function}
load_tcga_brca_data <- function() {
  # 1. Fetch data
  brca <- curatedTCGAData(
    diseaseCode = "BRCA",
    assays = c("RNASeq2GeneNorm", "Methylation*"),
    version = "2.1.1",
    dry.run = FALSE
  )

  # Helper to safe-extract an assay to a standard numeric matrix
  get_matrix <- function(mae, pattern) {
    found <- grep(pattern, names(assays(mae)), value = TRUE)
    if (length(found) == 0) stop(paste("No assay found for pattern:", pattern))
    mat <- as.matrix(assays(mae)[[found[1]]])
    if (!is.numeric(mat)) mode(mat) <- "numeric"
    return(mat)
  }

  rna <- get_matrix(brca, "RNASeq2GeneNorm")
  meth <- get_matrix(brca, "Methylation")

  # 2. Match patients by truncated barcode (first 12 characters)
  clean_barcode <- function(b) substr(b, 1, 12)
  rna_pats <- clean_barcode(colnames(rna))
  meth_pats <- clean_barcode(colnames(meth))
  common_pats <- intersect(rna_pats, meth_pats)
  if (length(common_pats) < 10) stop("Too few common patients. Matching failed.")

  rna <- rna[, match(common_pats, rna_pats)]
  meth <- meth[, match(common_pats, meth_pats)]
  colnames(rna) <- common_pats
  colnames(meth) <- common_pats

  # 3. Outcome from the clinical table
  col_data <- colData(brca)
  col_pats <- clean_barcode(rownames(col_data))
  matched_col_data <- col_data[match(common_pats, col_pats), ]
  y_raw <- matched_col_data$vital_status
  y <- ifelse(y_raw %in% c("deceased", "1", 1), 1, 0)
  y[is.na(y_raw)] <- NA

  keep_mask <- !is.na(y)
  rna <- rna[, keep_mask]
  meth <- meth[, keep_mask]
  y <- y[keep_mask]

  # 4. Clean and keep the top 1000 variable features per modality
  clean_and_filter <- function(mat, n_top = 1000) {
    na_counts <- rowSums(is.na(mat))
    mat <- mat[na_counts < ncol(mat), ]
    if (any(is.na(mat))) {
      row_means <- rowMeans(mat, na.rm = TRUE)
      na_idx <- which(is.na(mat), arr.ind = TRUE)
      mat[na_idx] <- row_means[na_idx[, 1]]
    }
    vars <- apply(mat, 1, var)
    keep_var <- !is.na(vars) & vars > 1e-6
    mat <- mat[keep_var, ]
    vars <- vars[keep_var]
    top_idx <- order(vars, decreasing = TRUE)[1:min(n_top, nrow(mat))]
    return(mat[top_idx, ])
  }

  X_rna <- t(clean_and_filter(rna, 1000))
  X_meth <- t(clean_and_filter(meth, 1000))
  colnames(X_rna) <- paste0("RNA_", colnames(X_rna))
  colnames(X_meth) <- paste0("METH_", colnames(X_meth))

  return(list(X_rna = X_rna, X_meth = X_meth, y = y))
}
```

Running it downloads the data, which takes a while, so the chatter is hidden. The summary below shows what came back.

```{r load-data-run, results='hide'}
tcga_data <- load_tcga_brca_data()
y <- tcga_data$y
X_rna <- tcga_data$X_rna
X_meth <- tcga_data$X_meth
```

```{r data-summary}
data_summary <- data.frame(
  Modality = c("RNA-seq", "Methylation"),
  Features = c(ncol(X_rna), ncol(X_meth)),
  Samples = c(nrow(X_rna), nrow(X_meth))
)
knitr::kable(data_summary, caption = "TCGA-BRCA analysis set after filtering")

outcome_tab <- data.frame(
  Status = c("Living (0)", "Deceased (1)"),
  N = as.numeric(table(y))
)
knitr::kable(outcome_tab, caption = "Outcome balance")
```

The adaptive weights. For each feature group I compute a group score (the L2 norm of the feature-outcome correlations), then a permutation p-value by shuffling the outcome 200 times. The penalty weight is 1 / (-log(p) + 1): groups with strong evidence get penalized less.

```{r adaptive-weights}
#' @title Calculate Adaptive Weights from Permutation P-values
calculate_p_value_weights_robust <- function(X, y, groups, n_perm = 200) {

  unique_groups <- levels(groups)
  y_centered <- y - mean(y)
  X_centered <- scale(X, center = TRUE, scale = FALSE)

  # Group score: L2 norm of the feature-outcome correlations
  calc_scores_fast <- function(y_vec, X_mat) {
    cor_vec <- as.numeric(crossprod(X_mat, y_vec))
    sapply(unique_groups, function(g) {
      idx <- which(groups == g)
      sqrt(sum(cor_vec[idx]^2))
    })
  }

  real_scores <- calc_scores_fast(y_centered, X_centered)

  null_scores <- matrix(NA, nrow = n_perm, ncol = length(unique_groups))
  for (k in 1:n_perm) {
    null_scores[k, ] <- calc_scores_fast(sample(y_centered), X_centered)
  }

  # Empirical p-values with a pseudo-count so p is never 0
  p_values <- sapply(seq_along(unique_groups), function(i) {
    (sum(null_scores[, i] >= real_scores[i]) + 1) / (n_perm + 1)
  })
  names(p_values) <- unique_groups

  # Inverse-log weighting: low p -> high surprisal -> low weight (low penalty)
  weights <- 1 / (-log(p_values) + 1)

  return(list(weights = weights, p_values = p_values))
}
```

Groups come from hierarchical clustering on 1 minus the absolute Spearman correlation, cut to about one group per ten features.

```{r define-groups}
define_feature_groups <- function(X) {
  cor_mat <- cor(X, method = "spearman")
  dist_mat <- as.dist(1 - abs(cor_mat))
  hclust_obj <- hclust(dist_mat, method = "ward.D2")
  n_groups <- max(1, round(ncol(X) / 10))
  groups <- cutree(hclust_obj, k = n_groups)
  return(as.factor(groups))
}
```

The pipeline for one feature matrix: standardize, define groups, compute adaptive weights, fit a cross-validated group lasso, and read off the selected features and the AUC at lambda.min.

```{r pipeline}
run_adaptive_lasso_pipeline <- function(X, y, n_perm = 200) {

  X <- scale(X) # z-score standardization
  groups <- define_feature_groups(X)

  weight_res <- calculate_p_value_weights_robust(X, y, groups, n_perm = n_perm)
  group_multipliers <- weight_res$weights[levels(groups)]

  cv_fit <- cv.grpreg(
    X, y,
    group = groups,
    penalty = "grLasso",
    family = "binomial",
    group.multiplier = group_multipliers,
    seed = 42
  )

  coefs <- coef(cv_fit, s = "lambda.min")
  selected <- names(coefs)[coefs != 0]
  selected <- selected[selected != "(Intercept)"]

  preds <- predict(cv_fit, X, s = "lambda.min", type = "response")
  auc_val <- as.numeric(auc(roc(y, as.vector(preds), quiet = TRUE)))

  return(list(
    selected_features = selected,
    auc = auc_val,
    weights = weight_res$weights,
    model = cv_fit
  ))
}
```

Phase 1: the three-way comparison. Early fusion concatenates both modalities and runs the pipeline once. Late fusion runs the pipeline on each modality separately and averages the two predicted probabilities. The baseline is a plain cross-validated lasso on the concatenated matrix.

```{r phase1-run, results='hide'}
X_early <- cbind(X_rna, X_meth)

early_res <- run_adaptive_lasso_pipeline(X_early, y, n_perm = 200)

late_rna_res <- run_adaptive_lasso_pipeline(X_rna, y, n_perm = 200)
late_meth_res <- run_adaptive_lasso_pipeline(X_meth, y, n_perm = 200)

pred_rna <- predict(late_rna_res$model, scale(X_rna), s = "lambda.min", type = "response")
pred_meth <- predict(late_meth_res$model, scale(X_meth), s = "lambda.min", type = "response")
late_prob <- (pred_rna + pred_meth) / 2
late_auc <- as.numeric(auc(roc(y, as.vector(late_prob), quiet = TRUE)))

cv_lasso <- cv.glmnet(scale(X_early), y, family = "binomial", alpha = 1)
lasso_prob <- predict(cv_lasso, scale(X_early), s = "lambda.min", type = "response")
lasso_auc <- as.numeric(auc(roc(y, as.vector(lasso_prob), quiet = TRUE)))

lasso_coefs <- coef(cv_lasso, s = "lambda.min")
lasso_selected <- rownames(lasso_coefs)[which(lasso_coefs != 0)]
lasso_selected <- lasso_selected[lasso_selected != "(Intercept)"]
lasso_count <- length(lasso_selected)
```

## Results

AUC and the number of selected features for each approach:

```{r phase1-table}
results_df <- data.frame(
  Model = c("Early Fusion", "Late Fusion", "Baseline Lasso"),
  AUC = c(early_res$auc, late_auc, lasso_auc),
  Features_Selected = c(
    length(early_res$selected_features),
    length(late_rna_res$selected_features) + length(late_meth_res$selected_features),
    lasso_count
  )
)
knitr::kable(results_df, digits = 3, caption = "TCGA-BRCA model comparison")
```

A sanity check that the adaptive weights did something sensible: the 40 feature blocks with the smallest weights (smallest weight means strongest evidence, so least penalty), colored by which modality their features come from.

```{r weights-plot, fig.width=7, fig.height=8}
weight_df <- data.frame(
  Block_ID = names(early_res$weights),
  Weight = as.numeric(early_res$weights)
)

groups_early <- define_feature_groups(scale(X_early))
feature_names <- colnames(X_early)

block_types <- sapply(names(early_res$weights), function(grp_id) {
  idx <- which(groups_early == grp_id)
  feats <- feature_names[idx]
  if (all(grepl("RNA", feats))) return("RNA block")
  if (all(grepl("METH", feats))) return("Methylation block")
  return("Mixed block")
})
weight_df$Type <- block_types

top_blocks <- weight_df %>%
  arrange(Weight) %>%
  head(40)

ggplot(top_blocks, aes(x = reorder(Block_ID, -Weight), y = Weight, fill = Type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("RNA block" = "steelblue",
                               "Methylation block" = "orange",
                               "Mixed block" = "gray")) +
  labs(
    title = "Top 40 most significant feature blocks (TCGA-BRCA)",
    subtitle = "Lower bar means stronger evidence, so less penalty",
    y = "Adaptive penalty weight",
    x = "Feature block ID"
  ) +
  theme_minimal()
```

## Method, continued: the stability experiment

AUC on one fit is not enough. I want to know whether the selected features stay the same when the data shifts. So I rerun the selection on five 80 percent subsamples and measure how much the selected sets overlap, using the Jaccard index between consecutive runs. This phase uses the knockoff version of the permutation-assisted group lasso: decoy features are made by permuting rows, and the lambda picked is the one where real selections beat decoy selections by the most.

```{r stability-helpers}
calculate_jaccard <- function(set1, set2) {
  union_len <- length(union(set1, set2))
  if (union_len == 0) return(1)
  return(length(intersect(set1, set2)) / union_len)
}

calculate_subsequent_jaccard <- function(feature_list) {
  n <- length(feature_list)
  if (n < 2) return(NA)
  scores <- sapply(1:(n - 1), function(i) calculate_jaccard(feature_list[[i]], feature_list[[i + 1]]))
  return(mean(scores))
}

#' @title Permutation-Assisted Group Lasso (knockoff selection)
perm_assisted_grplasso_binary <- function(X, y, n_perm, feature_groups, group_penalties, cutoff) {
  X <- as.matrix(X)
  p <- ncol(X)
  all_feature_names <- colnames(X)
  selected_features_all_perms <- c()

  g_int <- as.integer(feature_groups)
  knockoff_groups <- g_int + max(g_int)
  # Knockoffs inherit the exact same penalty as their original group
  full_group_penalties <- c(group_penalties, group_penalties)

  for (k in 1:n_perm) {
    X_knockoff <- X[sample(nrow(X)), ]

    fit <- suppressWarnings(grpreg(
      cbind(X, X_knockoff), y,
      group = c(g_int, knockoff_groups),
      penalty = "grLasso",
      family = "binomial",
      group.multiplier = full_group_penalties,
      nlambda = 50,
      warn = FALSE
    ))

    betas <- as.matrix(fit$beta[-1, ])

    n_original <- colSums(abs(betas[1:p, ]) > 0)
    n_knockoff <- colSums(abs(betas[(p + 1):(2 * p), ]) > 0)
    diff <- n_original - n_knockoff

    optimal_idx <- if (all(diff <= 0)) NULL else which.max(diff)

    if (!is.null(optimal_idx)) {
      selected_indices <- which(abs(betas[1:p, optimal_idx]) > 0)
      if (length(selected_indices) > 0) {
        selected_features_all_perms <- c(selected_features_all_perms, rownames(betas)[selected_indices])
      }
    }
  }

  freq_counts <- setNames(rep(0, p), all_feature_names)
  if (length(selected_features_all_perms) > 0) {
    counts_table <- table(selected_features_all_perms)
    freq_counts[names(counts_table)] <- counts_table
  }

  feature_frequencies <- freq_counts / n_perm
  final_selection <- names(feature_frequencies)[feature_frequencies >= cutoff]

  return(list(final_selection = final_selection))
}

run_stability_experiment <- function(mode) {

  feature_lists <- list()
  auc_scores <- numeric(N_RUNS)

  for (i in 1:N_RUNS) {
    set.seed(i)
    # 80% subsample
    train_idx <- sample(seq_len(nrow(X_early)), floor(0.8 * nrow(X_early)))
    y_sub <- y[train_idx]

    if (mode == "Early Fusion") {
      X_sub <- X_early[train_idx, ]

      res <- perm_assisted_grplasso_binary(
        X = X_sub, y = y_sub,
        n_perm = N_PERM_SELECT,
        feature_groups = groups_early,
        group_penalties = weights_early[levels(groups_early)],
        cutoff = 0.5
      )

      sel <- res$final_selection
      feature_lists[[i]] <- sel

      # AUC from a ridge refit on the selected features
      if (length(sel) > 1) {
        fit_eval <- cv.glmnet(scale(X_sub[, sel]), y_sub, family = "binomial", alpha = 0)
        p <- predict(fit_eval, scale(X_sub[, sel]), s = "lambda.min", type = "response")
        auc_scores[i] <- as.numeric(auc(roc(y_sub, as.vector(p), quiet = TRUE)))
      } else { auc_scores[i] <- 0.5 }

    } else if (mode == "Late Fusion") {
      res_rna <- perm_assisted_grplasso_binary(
        X = X_rna[train_idx, ], y = y_sub,
        n_perm = N_PERM_SELECT,
        feature_groups = groups_rna,
        group_penalties = weights_rna[levels(groups_rna)],
        cutoff = 0.5
      )

      res_meth <- perm_assisted_grplasso_binary(
        X = X_meth[train_idx, ], y = y_sub,
        n_perm = N_PERM_SELECT,
        feature_groups = groups_meth,
        group_penalties = weights_meth[levels(groups_meth)],
        cutoff = 0.5
      )

      sel <- unique(c(res_rna$final_selection, res_meth$final_selection))
      feature_lists[[i]] <- sel

      # AUC from the averaged ensemble probabilities
      auc_rna <- 0.5; auc_meth <- 0.5

      if (length(res_rna$final_selection) > 0) {
         fA <- cv.glmnet(scale(X_rna[train_idx, res_rna$final_selection]), y_sub, family = "binomial", alpha = 0)
         probs_rna <- predict(fA, scale(X_rna[train_idx, res_rna$final_selection]), s = "lambda.min", type = "response")
         auc_rna <- as.numeric(auc(roc(y_sub, as.vector(probs_rna), quiet = TRUE)))
      }

      if (length(res_meth$final_selection) > 0) {
         fB <- cv.glmnet(scale(X_meth[train_idx, res_meth$final_selection]), y_sub, family = "binomial", alpha = 0)
         probs_meth <- predict(fB, scale(X_meth[train_idx, res_meth$final_selection]), s = "lambda.min", type = "response")
         auc_meth <- as.numeric(auc(roc(y_sub, as.vector(probs_meth), quiet = TRUE)))
      }

      auc_scores[i] <- (auc_rna + auc_meth) / 2

    } else if (mode == "Baseline Lasso") {
      fit <- cv.glmnet(scale(X_early[train_idx, ]), y_sub, family = "binomial", alpha = 1)

      coefs <- coef(fit, s = "lambda.min")
      sel <- rownames(coefs)[which(coefs != 0)]
      sel <- sel[sel != "(Intercept)"]
      feature_lists[[i]] <- sel

      p <- predict(fit, scale(X_early[train_idx, ]), s = "lambda.min", type = "response")
      auc_scores[i] <- as.numeric(auc(roc(y_sub, as.vector(p), quiet = TRUE)))
    }
  }

  return(list(
    mode = mode,
    stability = calculate_subsequent_jaccard(feature_lists),
    mean_auc = mean(auc_scores),
    sd_auc = sd(auc_scores),
    avg_features = mean(sapply(feature_lists, length))
  ))
}
```

The global weights are computed once up front with 500 permutations, then reused inside each subsample run. Five runs, fifty inner permutations per selection, 0.5 frequency cutoff.

```{r stability-run, results='hide'}
# Global weights (computed once, reused across subsamples)
groups_early <- define_feature_groups(X_early)
weights_early <- calculate_p_value_weights_robust(X_early, y, groups_early, n_perm = 500)$weights

groups_rna <- define_feature_groups(X_rna)
weights_rna <- calculate_p_value_weights_robust(X_rna, y, groups_rna, n_perm = 500)$weights

groups_meth <- define_feature_groups(X_meth)
weights_meth <- calculate_p_value_weights_robust(X_meth, y, groups_meth, n_perm = 500)$weights

N_RUNS <- 5
N_PERM_SELECT <- 50

res_early <- run_stability_experiment("Early Fusion")
res_late <- run_stability_experiment("Late Fusion")
res_baseline <- run_stability_experiment("Baseline Lasso")
```

## Results, continued

Stability (mean Jaccard overlap between consecutive runs), mean AUC, and the average number of selected features:

```{r stability-table}
stability_df <- data.frame(
  Method = c("Early Fusion (Adaptive)", "Late Fusion (Adaptive)", "Baseline Lasso"),
  Stability_Jaccard = c(res_early$stability, res_late$stability, res_baseline$stability),
  Mean_AUC = c(res_early$mean_auc, res_late$mean_auc, res_baseline$mean_auc),
  Avg_Features = c(res_early$avg_features, res_late$avg_features, res_baseline$avg_features)
)
knitr::kable(stability_df, digits = 3,
             caption = "Stability and performance over 5 subsample runs (TCGA-BRCA)")
```

```{r stability-plot, fig.width=8, fig.height=4}
plot_df <- stability_df %>%
  pivot_longer(cols = c(Stability_Jaccard, Mean_AUC), names_to = "Metric", values_to = "Value")

ggplot(plot_df, aes(x = Method, y = Value, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.9) +
  facet_wrap(~Metric, scales = "free_y") +
  theme_bw() +
  scale_fill_manual(values = c("gray40", "forestgreen", "steelblue")) +
  labs(
    title = "Performance versus stability",
    subtitle = "Adaptive fusion methods against the baseline",
    y = "Score"
  ) +
  theme(legend.position = "bottom")
```

## Takeaway

The two tables answer the two questions. The first table shows whether the adaptive fusion methods compete with plain lasso on prediction while selecting fewer, more interpretable features. The second table is the stricter test: a method that picks a different feature set every time the data shifts slightly is not one to trust, no matter its AUC. If early fusion holds both its AUC and its Jaccard stability above the baseline here, it is the method worth carrying into the thesis.
